---
title: "Getting Started with vastR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with vastR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vastR)
```

## What is VAST?

The **Visual Argument Structure Tool** (VAST) was introduced by
Leising, Grenke, & Cramer (2023) as a system for visually mapping the
structure of arguments. It distinguishes between:

- **Concepts** and their **names** (words)
- Multiple relationship types: causation, implication, prediction,
  transformation, reasoning
- **IS** and **OUGHT** statements
- **Perspectives** of different people on those statements
- **Higher-order concepts** that group elements together

The `vastR` package lets you build VAST diagrams programmatically in R.

## A First Diagram

Let's recreate the classic Sun/Hot/Bright example from the paper:

```{r basic-example}
model <- vast_model(title = "Conceptual Implications") |>
  add_nodes(
    vast_concept("S", "S"),
    vast_concept("H", "H"),
    vast_concept("B", "B"),
    vast_name("n_sun",    "Sun"),
    vast_name("n_hot",    "Hot"),
    vast_name("n_bright", "Bright")
  ) |>
  add_edges(
    vast_implication("S", "H"),
    vast_implication("S", "B"),
    vast_naming("S", "n_sun"),
    vast_naming("H", "n_hot"),
    vast_naming("B", "n_bright")
  )

vast_render(model)
```

## Naming Modes

VAST distinguishes concepts (abstract cognitive categories) from names
(the words used to refer to them). The `naming_mode` parameter on
`vast_model()` controls how this distinction is displayed. You always
define both concepts and name nodes with naming edges — the mode only
changes the rendering:

**Integrated** (default): Concepts and names coexist in the diagram,
connected by naming (n) arrows.

```{r naming-integrated, eval=FALSE}
vast_model(naming_mode = "integrated")  # this is the default
```

**Separated**: The main diagram shows only concepts with abstract labels.
A separate legend box displays the concept-to-name mappings.

```{r naming-separated, eval=FALSE}
vast_model(naming_mode = "separated")
```

**FIMM** (Finger-is-Moon-Mode): Concept nodes display the label of their
first connected name node directly. Name nodes and naming edges are
hidden.

```{r naming-fimm, eval=FALSE}
vast_model(naming_mode = "fimm")
```

You can switch between modes by changing a single parameter — the model
definition stays the same. See Example 1 (`inst/examples/01_conceptual_implication.R`)
for a full demonstration.

## Relationship Types

VAST distinguishes six core relationship types. Each has a dedicated
constructor with its own default color:

```{r relationship-types}
model <- vast_model(title = "Relationship Types", naming_mode = "fimm", rankdir = "LR") |>
  add_nodes(
    vast_concept("Smoking",    "Smoking"),
    vast_concept("LungCancer", "Lung Cancer"),
    vast_concept("Height",     "Height"),
    vast_concept("YChrom",     "Y Chromosomes")
  ) |>
  add_edges(
    vast_causation("Smoking", "LungCancer"),
    vast_prediction("Height", "YChrom")
  )

vast_render(model)
```

## Relationship Strength

All edge functions accept a `strength` parameter — numeric, verbal, or
special markers:

```{r strengths, eval=FALSE}
vast_causation("X", "Y", strength = 0.5)      # Numeric
vast_causation("X", "Y", strength = "weak")    # Verbal
vast_causation("X", "Y", strength = "?")       # Unknown
vast_causation("X", "Y", strength = "<> 0")    # Nonzero
```

## Diamonds: Logical Connectives and Custom Functions

Diamonds can hold any label — standard connectives or formulas:

```{r diamonds}
model <- vast_model(naming_mode = "fimm", rankdir = "LR") |>
  add_nodes(
    vast_concept("X", "X"),
    vast_concept("Z", "Z"),
    vast_concept("Y", "Y"),
    vast_diamond("fn", "Y = 2.45X + Z")
  ) |>
  add_edges(
    vast_causation("X", "fn"),
    vast_causation("Z", "fn"),
    vast_causation("fn", "Y")
  )

vast_render(model)
```

## Higher-Order Concepts (Groups)

Groups draw a bounding box around related elements. They can also be
nested:

```{r groups}
model <- vast_model(title = "Nested Groups", naming_mode = "fimm") |>
  add_nodes(
    vast_concept("A", "A"),
    vast_concept("B", "B"),
    vast_concept("C", "C"),
    vast_concept("D", "D"),
    vast_diamond("d1", "AND")
  ) |>
  add_edges(
    vast_causation("A", "d1"),
    vast_causation("B", "d1"),
    vast_causation("C", "d1"),
    vast_causation("d1", "D")
  ) |>
  add_groups(
    vast_group("g1", label = "Group 1", node_ids = c("A", "B"),
               fillcolor = "#FCE4EC", bordercolor = "#C62828"),
    vast_group("g2", label = "Group 2", node_ids = c("C"),
               fillcolor = "#E8F5E9", bordercolor = "#2E7D32"),
    vast_nested_group("outer", label = "All Inputs",
                      node_ids = c("d1"),
                      child_group_ids = c("g1", "g2"))
  )

vast_render(model)
```

## IS, OUGHT, and Perspectives

```{r is-ought}
model <- vast_model(title = "Beliefs and Perspectives") |>
  add_nodes(
    vast_concept("X", "X"),
    vast_is("is1", value = "0.7"),
    vast_ought("o1", value = "1.0"),
    vast_perspective("p1", holder = "Alice", value = "0.9"),
    vast_perspective("p2", holder = "Bob",   value = "0.4")
  ) |>
  add_edges(
    vast_relation("X", "is1", type = "", color = "gray50"),
    vast_relation("X", "o1",  type = "", color = "gray50"),
    vast_relation("p1", "is1", type = "", color = "gray50", style = "dashed"),
    vast_relation("p2", "is1", type = "", color = "gray50", style = "dashed")
  )

vast_render(model)
```

## Exporting

```{r export, eval=FALSE}
vast_export_svg(model, "my_diagram.svg")
vast_export_png(model, "my_diagram.png", width = 1200, height = 800)
```

## Inspecting the DOT Code

If you want to see or tweak the raw Graphviz DOT code:

```{r dot-output}
cat(vast_to_dot(model))
```

## Reference

Leising, D., Grenke, O., & Cramer, M. (2023). Visual Argument Structure
Tool (VAST) Version 1.0. *Meta-Psychology*, 7.
doi: [10.15626/MP.2021.2911](https://doi.org/10.15626/MP.2021.2911)
